---
title: "Taylor Swift Song and Lyric Analysis"
author: "Dingning Yang"
date: "2025-02-15"
format: 
 html:
    toc: true
    toc-depth: 3
    toc-title: "Table of Contents"
    number-sections: true
    highlight-style: github
    theme: cosmo
    fontsize: 1.25em
    linestretch: 1.5
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
# Install required packages
if(!require(tidyverse)){install.packages("tidyverse")}
if(!require(ggplot2)){install.packages("ggplot2")}
if(!require(ggiraph)){install.packages("ggiraph")}
if(!require(viridis)){install.packages("viridis")}
if(!require(knitr)){install.packages("knitr")}
if(!require(kableExtra)){install.packages("kableExtra")}
if(!require(plotly)){install.packages("plotly")}
if(!require(wordcloud2)){install.packages("wordcloud2")}
if(!require(shiny)){install.packages("shiny")}

# List of required packages
required_packages <- c ("tidyverse", "ggplot2", "ggiraph", "viridis", "knitr", "kableExtra", "plotly", "wordcloud2", "shiny")

# Load required packages
lapply(required_packages, library, character.only = TRUE)
set.seed(123)

```

## Spotify Data Analysis

This project first focuses on analyzing Taylor Swift's music using data from her songs' Spotify data. The raw dataset from Kaggle includes different data types such as album name, track name, release date (***categorical***); acousticness, danceability, energy, popularity scores (***numeric***), etc. Data cleaning and transformation were performed before analyzing the data.

```{r}
#| label: Source cleaned and transformed sourced script
#| include: false

source("Source/Data Cleaning.R")

```

### Descriptive Analysis

The Spotify dataset contains `r n_distinct (ts.spotify$id)` tracks from `r n_distinct (ts.spotify$album)` albums of Taylor Swift. Specifically, there are `r nrow(ts.spotify[ts.spotify$album_label == "Republic", ])` tracks and `r length(unique(ts.spotify$album[ts.spotify$album_label == "Republic"]))` albums under the Republic record label. Under the Big Machine record label, there are `r nrow(ts.spotify[ts.spotify$album_label == "Big Machine", ])` tracks and `r length(unique(ts.spotify$album[ts.spotify$album_label == "Big Machine"]))` albums. Table 1 summarizes the album release time and average popularity scores of different albums.

```{r}
#| label: Generate album release time and popularity table
#| tbl-cap: "Table 1: Album Release Time and Popularity"

album_summary <- ts.spotify %>%
  mutate(track_number = as.numeric(as.character(track_number))) %>% 
  group_by(album) %>%
  summarise(
    album_label = first(album_label),
    release_year = first(release_year),
    release_month = first(release_month),
    avg_popularity = sum(popularity) / max(track_number),
    .groups = 'drop'
  ) %>%
  arrange(desc(release_year), desc(release_month), desc(avg_popularity))
  
# Create the kable table
album_summary %>%
  kable(
    caption = "Album Release Time and Popularity",
    col.names = c("Album", "Label", "Year", "Month", "Avg. Popularity"),
    digits = 2,
    align = c('l', 'l', 'c', 'c', 'r')
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(0, bold = TRUE)




```

Specifically, `r round(100 * sum(ts.spotify$album_label == "Republic") / nrow(ts.spotify), 1)`% of Taylor Swift's songs are under the Republic record label, while `r round(100 * sum(ts.spotify$album_label == "Big Machine") / nrow(ts.spotify), 1)`% are under the Big Machine record label.

```{r}
#| label: fig-album-label-percentage
#| fig.cap: "Percentage of Taylor Swift's Songs by Record Label"
#| include: true

# Summarize data for album label
label_data <- ts.spotify %>%
  group_by(album_label) %>%
  summarise(count = n(), .groups = 'drop') %>%
  mutate(percentage = count / sum(count) * 100)  # Calculate percentage

# Create the pie chart
p <- plot_ly(label_data, 
              labels = ~album_label, 
              values = ~count, 
              type = 'pie', 
              textinfo = 'label+percent', 
              hoverinfo = 'text', 
              text = ~paste(album_label, ": ", count, " (", round(percentage, 1), "%)", sep = ""),
              marker = list(colors = viridis::viridis(nrow(label_data)))) %>%
  layout(title = "Percentage of Taylor Swift's Songs by Record Label",
         showlegend = TRUE,
         legend = list(orientation = "h", xanchor = "center", x = 0.5, y = -0.2))

# Show the plot
p

```


The dataset also includes music features of each song, which are explained in the following sections:

#### Acousticness

A confidence measure from 0.0 to 1.0 of whether the track is acoustic. Tracks with values close to 0.0 are likely to have minimal acoustic elements and are more likely produced using electronic instruments or heavily synthesized sounds. Tracks with values close to 1.0 are likely to have a strong presence of acoustic elements, such as acoustic guitar, piano, or other non-electronic instruments.

For instance, the lowest acousticness score is 0.000 for ***Style*** and the highest is 0.971 for ***It’s Nice To Have A Friend***.

##### Style

Click the button below to listen to ***Style***: Low Acousticness Score

<audio controls>

<source src="audio/Taylor Swift - Style.mp3" type="audio/mpeg">

Your browser does not support the audio element. </audio>

##### It's Nice To Have A Friend

Click the button below to listen to ***It's Nice To Have A Friend***: High Acousticness Score

<audio controls>

<source src="audio/Taylor Swift - It's Nice To Have A Friend.mp3" type="audio/mpeg">

Your browser does not support the audio element. </audio>

##### Summary

Specifically, the average acousticness score of Taylor Swift's songs is `r round(mean(ts.spotify$acousticness), 2)`(*SD* = `r round(sd(ts.spotify$acousticness, na.rm = TRUE), 1)`).

#### Danceability

Danceability describes how suitable a track is for dancing based on a combination of musical elements including tempo, rhythm stability, beat strength, and overall regularity. A value of 0.0 is least danceable and 1.0 is most danceable.

For instance, the lowest danceability score is 0.175 for ***epiphany*** and the highest is 0.897 for ***I Think He Knows***.

##### epiphany

Click the button below to listen to ***epiphany***: Low Danceability Score

<audio controls>

<source src="audio/Taylor Swift - epiphany.mp3" type="audio/mpeg">

Your browser does not support the audio element. </audio>

##### I Think He Knows

Click the button below to listen to ***I Think He Knows***: High Danceability Score

<audio controls>

<source src="audio/Taylor Swift - I Think He Knows.mp3" type="audio/mpeg">

Your browser does not support the audio element. </audio>

##### Summary
Specifically, the average danceability score of Taylor Swift's songs is `r round(mean(ts.spotify$danceability), 2)`(*SD* = `r round(sd(ts.spotify$danceability, na.rm = TRUE), 1)`).


#### Energy

Energy is a measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. Typically, energetic tracks feel fast, loud, and noisy.

For instance, the lowest energy score is 0.118 for ***State Of Grace - Acoustic*** and the highest is 0.948 for ***Haunted***.

##### State Of Grace - Acoustic

Click the button below to listen to ***State Of Grace - Acoustic***: Low Energy Score

<audio controls>

<source src="audio/Taylor Swift - State Of Grace (Acoustic Version) (Taylor's Version).mp3" type="audio/mpeg">

Your browser does not support the audio element. </audio>

##### Haunted

Click the button below to listen to ***Haunted***: High Energy Score

<audio controls>

<source src="audio/Taylor Swift - Haunted.mp3" type="audio/mpeg">

Your browser does not support the audio element. </audio>

##### Summary
Specifically, the average energy score of Taylor Swift's songs is `r round(mean(ts.spotify$energy), 2)`(*SD* = `r round(sd(ts.spotify$energy, na.rm = TRUE), 1)`).


#### Liveness

Liveness detects the presence of an audience in the recording. Higher liveness values represent an increased probability that the track was performed live. A value above 0.8 provides strong likelihood that the track is live.

For instance, the lowest liveness score is 0.034 for ***I Knew You Were Trouble*** and the highest is 0.931 for ***Better Than Revenge - Live/2011***.

##### I Knew You Were Trouble

Click the button below to listen to ***I Knew You Were Trouble***: Low Liveness Score

<audio controls>

<source src="audio/Taylor Swift - I Knew You Were Trouble (Taylor's Version).mp3" type="audio/mpeg">


##### Better Than Revenge - Live/2011

Click the button below to listen to ***Better Than Revenge - Live/2011***: High Liveness Score

<audio controls>

<source src="audio/Taylor Swift - Better Than Revenge (Live2011).mp3" type="audio/mpeg">

##### Summary

Specifically, the average liveness score of Taylor Swift's songs is `r round(mean(ts.spotify$liveness), 2)`(*SD* = `r round(sd(ts.spotify$liveness, na.rm = TRUE), 1)`). It is notable that most albums have a liveness score close to 0, indicating that the majority of the songs are studio albums rather than live recordings. Nevertheless, it is possible that the same song may have different versions, such as live versions, acoustic versions, or remixes, which may have higher liveness scores. For instance, ***I Knew You Were Trouble*** has multiple live versions. Here is an example from reputation Stadium tour (Oct 19, 2018):

<video width="320" height="240" controls>
     <source src="audio/IKYWT live2018.mp4" type="video/mp4">
     Your browser does not support the video tag.
   </video>
   
   
#### Valence

A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry).

For instance, the lowest valence score is 0.153 for ***Maroon*** and the highest is 0.943 for ***Shake It Off***.

##### Maroon

Click the button below to listen to ***Maroon***: Low Valence Score

<audio controls>

<source src="audio/Taylor Swift - Maroon.mp3" type="audio/mpeg">



##### Shake It Off

Click the button below to listen to ***Shake It Off***: High Valence Score

<audio controls>

<source src="audio/Taylor Swift - Shake It Off.mp3" type="audio/mpeg">

##### Summary

Specifically, the average valence score of Taylor Swift's songs is `r round(mean(ts.spotify$valence), 2)`(*SD* = `r round(sd(ts.spotify$valence, na.rm = TRUE), 1)`). This indicates that her songs are slightly skewed towards negative emotions. 


It is reasonable to conclude that the song ***Maroon*** has a low valence score based on the the lyrics and the muse of the song. Specifically, the lyrics are proven to be related to several plots in a book called "*The Seven Husbands of Evelyn Hugo*" by Taylor Jenkins Reid. The novel follows Evelyn rose from poverty to become a major star in the 1950s-60s, navigating the complicated world of Hollywood while hiding her true identity as a bisexual Cuban woman by seven marriages. "*The rust that grew between telephone*" is a metaphor for the distance between Evelyn and Celia when they experienced public love blackouts, and the "maroon" color is a symbol of both the joy they shared together (drinking wine) and pain that Evelyn has experienced. The book is claimed to be paralleled with Taylor Swift's life about her beard romantic relationships and love blackout with the New York muse Karlie Kloss. Karlie Kloss confirmed her identity as the ***Maroon*** muse by wearing the "Cheap Ass Rosé" bracelet to the Eras tour on Aug 9, 2023 in Los Angeles, which resembles the "roommate muse" in ***Maroon***. 

It is also easy to tell ***Shake It Off***  is a positive song if you listen to it!


***Note***: Loudness is not discussed here since all values are negative. Speechiness is not discussed here since all values are close to zero. Tempo is not considered in the popularity score prediction since it is unrelated. 

### Popularity Score Predictors

#### Record Labels

Taylor Swift left Big Machine Records, her original label since age 15, and signed with Universal Music Group's Republic Records in November 2018. This move came after her contract with Big Machine expired. Taylor wanted to own her master recordings, which Big Machine wouldn't allow. After leaving, Big Machine sold her back catalog to Scooter Braun's company, leading to a public dispute. Unable to buy back her masters, Taylor began re-recording her first six albums (i.e., Taylor's Version) to regain control of her work. This section presents the popularity scores of Taylor Swift's songs under different record labels, aiming to see if her control over her work affects the popularity of her songs.

```{r}
#| label: ANOVA for popularity scores by record label
#| include: false

# Perform ANOVA for popularity scores by record label
anova_pop_label <- aov(popularity ~ album_label, data = ts.spotify)
summary (anova_pop_label)
summary_label_effect <- summary(anova_pop_label)[[1]]

# Check effect size
SSbetween <- summary_label_effect$`Sum Sq`[1]  # Sum of squares for the factor
SStotal <- sum(summary_label_effect$`Sum Sq`)  # Total sum of squares

eta_squared <- SSbetween / SStotal

# Round to 3 decimal places
eta_squared_rounded <- round(eta_squared, 3)


```

One-way ANOVA results indicated significant differences were found between record labels' impacts on popularity scores (Figure 2). Specifically, songs under the Big Machine label showed lower popularity scores than under the Republic label ($M_{\text{Big Machine}} = `r round(mean(ts.spotify$popularity[ts.spotify$album_label == "Big Machine"]), 2)`$ vs. $M_{\text{Republic}} = `r round(mean(ts.spotify$popularity[ts.spotify$album_label == "Republic"]), 2)`$; *F*(`r summary_label_effect$Df[1]`, `r summary_label_effect$Df[2]`) = `r round(summary_label_effect$"F value"[1], 2)`; *p* `r if(summary_label_effect$"Pr(>F)"[1] < 0.001) "< 0.001" else paste("=", sprintf("%.3f", summary_label_effect$"Pr(>F)"[1]))`, η² = `r eta_squared_rounded`). It indicates that approximately 46.5% of the variance in popularity scores can be explained by the difference in record labels (Big Machine vs. Republic), suggesting that Taylor Swift's control over her work has a significant impact on the popularity of her songs. The ownership not only allows her to re-record her songs but also to control the distribution and promotion of her music, which may contribute to the higher popularity scores under the Republic label. It is notable that the re-record versions are generally promoted by fans as a way to support Taylor Swift's ownership of her work. This also reflects Taylor's determination to regain control of her work and the revolutionary impact of her move on the music industry.




```{r}
#| label: fig-popularity-by-label
#| fig.cap: "Popularity Scores by Record Label"

# Define the theme
theme.clean <- function(){
  theme_bw()+
    theme(axis.text.x = element_text(size = 12, vjust = 1, hjust = 1),
          axis.text.y = element_text(size = 12),
          axis.title.x = element_text(size = 14, face = "plain"),             
          axis.title.y = element_text(size = 14, face = "plain"),
          panel.grid.major.x = element_blank(),                                          
          panel.grid.minor.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          panel.grid.major.y = element_blank(),  
          plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), units = , "cm"),
          plot.title = element_text(size = 16, vjust = 1, hjust = 0.5),
          plot.caption = element_text (size = 8,vjust = 1, hjust = 0.9),
          legend.text = element_text(size = 12, face = "italic"),          
          legend.position = "right")}

# Create the boxplot

label_effect_boxplot <- ggplot(ts.spotify, aes(album_label, popularity)) + 
  geom_boxplot(aes(fill = album_label)) +
  theme_bw() +
  scale_fill_manual(values = c("#f6ed95", "#800000")) +               
  scale_colour_manual(values = c("#f6ed95", "#800000")) + 
  labs(title = "Record Label Comparison of Song Popularity Impacts", 
       x = "Record Label", y = "Song Popularity Score") +
  theme.clean()+
  theme(axis.text.x = element_text(hjust = 0.5))+
  scale_y_continuous(breaks = seq(floor(min(ts.spotify$popularity)), 
                                  ceiling(max(ts.spotify$popularity)), by = 10)) +
  stat_summary(fun = mean, geom = "point", color = "black", size = 3, shape = 18, 
               position = position_dodge(0.75), aes(shape = "Mean"), 
               show.legend = FALSE) +  
  stat_summary(fun = mean, geom = "text", aes(label = round(after_stat(y), 2)), 
               color = "white", vjust = -1, position = position_dodge(0.75), 
               size = 3.5, show.legend = FALSE) +
  labs(fill = "Record Label") 

print (label_effect_boxplot)


```

## Running Code


